---
title: "Indicator_value_analysis"
author: "Patrick Pata"
date: "27/10/2021"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load files and libraries
```{r}
ls.packages <- c("tidyverse", "vegan","adespatial","ggrepel",
                 "tictoc","labdsv","here")
new.packages <- ls.packages[!(ls.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(tidyverse)
library(vegan)
library(adespatial)
library(ggrepel)
library(tictoc)
library(labdsv)
library(here)
`%notin%` <- Negate(`%in%`)

load(here::here("Data_Output/zoop_data_regionalized_10252021.RData"))

source(here::here("Functions/p_indvalScatter_25.R"))
```

Calculate IndVal using the Dufrene-Legendre method in labdsv package. 
Calculate the indicator value from the untransformed species abundance table although the results of this calculated on the log-transformed abundance is very similar (r ~ 0.95).
```{r, fig.width=10, fig.height=8}
indval.zoop <- indval(zoop_abundance, zoopmeta$Bioregion,
                      numitr=10000)
# indval.zoop <- indval(zoop.log, zoopmeta$Bioregion,
#                       numitr=10000)

summary(indval.zoop, p=0.05, type='short', digits=2, show=p,
          sort=FALSE, too.many=100)

sigsp25 <- p_indvalScatter_25(indval.zoop, plot = TRUE)
```


# Extract the indval table and the A and B components 
(A: Specificity/RelAbu x B: Fidelity/RelFrq)
```{r}
indval.zoop.bioregions <- data.frame(cluster = indval.zoop$maxcls, 
                                     indval = indval.zoop$indcls*100, 
                                     pval = indval.zoop$pval,
                                     padj = p.adjust(indval.zoop$pval, "holm")) %>% 
  rownames_to_column("Species") %>% 
  mutate(Bioregion = levels(bioregions$Bioregion)[cluster])
  
indval.zoop.AB <- data.frame(indval = indval.zoop$indval,
                              relabu = indval.zoop$relabu,
                             relfrq = indval.zoop$relfrq) %>% 
  bind_cols(dplyr::select(indval.zoop.bioregions, Species, Bioregion))
```

# Save indval results for relating with species associations.
```{r, eval =FALSE}
save(file = here::here(paste0("Data_Output/Indicator_values_11022021.RData")),
list = c("indval.zoop.bioregions","indval.zoop.AB"))
```


# NMDS of indicator value analysis results
```{r, fig.height=11, fig.width=11}
indval.nmds <- metaMDS(indval.zoop.AB[,c(5:12)])
(indval.nmds)
stressplot(indval.nmds)
plot(indval.nmds, display = c("sites","species", choices = c(1,2)) )
points(indval.nmds, display = c("sites", "species"),
    choices = c(1,2)) 

# NMDS scores for biplot
indval.nmds.scores <- as.data.frame( scores(indval.nmds) ) %>% 
  rownames_to_column("Species") %>% 
  left_join(indval.zoop.bioregions) %>% 
  mutate(Bioregion = as.factor(Bioregion))
indval.nmds.scores$Bioregion <- fct_relevel(indval.nmds.scores$Bioregion, 
                                            levels = c("Offshore","DeepShelf",
                                                       "Nearshore","DeepFjord"))

# Add to biplot the A & B components per cluster
AB.fit <- envfit(indval.nmds, indval.zoop.AB[,c(5:12)], 
                 permutations = 1000, strata = NULL,  choices=c(1,2), na.rm = TRUE)
AB.vec <- AB.fit$vectors$arrows * sqrt(AB.fit$vectors$r)
AB.vec <- as_tibble(AB.vec) %>%  
  mutate(labs = c("SPC-OS","SPC-DS","SPC-NS","SPC-DF",
                  "FID-OS","FID-DS","FID-NS","FID-DF") )


ggplot(indval.nmds.scores, 
       aes(x = NMDS1, y = NMDS2, color = Bioregion)) +
  geom_point() +
  scale_color_manual(values = clrs.bioregion) +
  theme_bw() +
  geom_text_repel(data = filter(indval.nmds.scores, indval >= 25, padj < 0.05),
                  aes(x = NMDS1, y = NMDS2, label = Species), fontface = "italic",
                  size = 4, 
                  max.overlaps = 21) +
  geom_segment(data = AB.vec, aes(x = rep(0,nrow(AB.vec)), y =rep(0,nrow(AB.vec)), 
                              xend = NMDS1,yend = NMDS2), 
               lineend = "round", size = 1, colour = "black") +
    geom_text_repel(data = AB.vec, aes(x = NMDS1, y = NMDS2, label=labs), 
                  color="black", fontface = 'bold', size = 6) +
  theme(legend.position = "none")
```

Alternative indval method using the indicspecies package of De Caceres.
```{r, eval = FALSE}
# library(indicspecies) # alternative to indval package

indspe.zoop <- multipatt(zoop_abundance, zoopmeta$Bioregion, duleg = TRUE)

indspe.zoop.table <- cbind(indval = as.data.frame(indspe.zoop$str ^ 2) * 100, 
              A = indspe.zoop$A* 100, B = indspe.zoop$B* 100,
              clust = indspe.zoop$sign$index, iv = (indspe.zoop$sign$stat^2) * 100, 
              pval = indspe.zoop$sign$p.value) %>% 
  rownames_to_column(var = "Taxa")

# # If analyzing with duleg = FALSE, a table of all combinations of clusters is returned
# indspe.zoop <- multipatt(spe.sub, clusters.k4, duleg = FALSE)
```

