---
title: "Species_association"
author: "Patrick Pata"
date: "27/10/2021"
output: html_document
---


Performs the concordance analysis for Table 3 and Table S2 for Pata et al. PinO manuscript.

For the species association analysis, a concordance analysis is done with the assumption that species association are hierarchically structured. The log-transformed abundances are correlated between species and the correlation matrix is then converted to a distance matrix for clustering. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load files and libraries
```{r}
ls.packages <- c("tidyverse","vegan","ggrepel","tictoc","factoextra",
                 "dendextend","Polychrome","fpc","here")
new.packages <- ls.packages[!(ls.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(tidyverse)
library(vegan)
library(ggrepel)
library(tictoc)
library(factoextra)
library(dendextend)
library(Polychrome)
library(fpc)
library(here)
`%notin%` <- Negate(`%in%`)

load(here::here("Data_Output/zoop_data_regionalized_10252021.RData"))
load(here::here("Data_Output/Indicator_values_11012021.RData"))

```

# Correlate and cluster
This follows steps 1-4 of the R-mode process in the kendall.global() description which is in Legendre (2012 p.395). Recommended in Borcard (2018, p. 52) in reference to Legendre (2005) that species data should be pre-transformed (box-cox series) to remove effect of total abundance per site in calculating correlations.
```{r, fig.width=12, fig.height=6}
# log-chord transformed table
assoc.cor <- cor(zoop.logchord, method = "spearman")
assoc.dist <- as.dist(1-assoc.cor)

assoc.clust <- hclust(assoc.dist, method = "ward.D2")
plot(assoc.clust, hang = -1, cex = 0.6)
```

# Calinski-Harabaz criterion from the species table for the number of clusters
```{r, eval=FALSE}
ch <- c()
for(k in seq(1,15)){
  clust.spe.cut <- cutree(assoc.clust, k = k) 
  ch[k] <- calinhara(t(zoop.logchord),clust.spe.cut, k)
}
plot(ch)
```

# Identify the number of association groups
```{r, fig.width=10, fig.height=5}
ceval <- matrix(NA, nrow = 30, ncol = 8)
for(k in seq(2,30)){
  clust.spe.cut <- cutree(assoc.clust, k = k) 
  cstat <- cluster.stats(assoc.dist, clust.spe.cut)
  
  ceval[k,1] <- cstat$avg.silwidth
  ceval[k,2] <- cstat$ch
  ceval[k,3] <- cstat$dunn
  ceval[k,4] <- cstat$wb.ratio
  ceval[k,5] <- cstat$pearsongamma  
  ceval[k,6] <- cstat$sindex
  ceval[k,7] <- cstat$entropy 
  ceval[k,8] <- cstat$within.cluster.ss
}

ceval <- as.data.frame(ceval[2:30,])
colnames(ceval) <- c("Avg.Silhouette.Width","Calinski.Harabaz","Dunn.Index",
                     "Within.Between.Ratio","Pearson.Gamma","Separation.Index",
                     "Entropy","Within.Sum.Squares")
ceval <- mutate(ceval, NumClusters = c(2:30))

ceval.long <- pivot_longer(ceval, -NumClusters, names_to = "Index", values_to = "Values")


plot(assoc.clust, hang = -1, cex = 0.6)

ggplot(ceval.long, aes(x = NumClusters, y = Values)) +
  geom_point() + geom_line() +
  theme_bw() +
  facet_wrap(~Index, scales = "free")

ggplot(filter(ceval.long, Index == "Avg.Silhouette.Width"), aes(x = NumClusters, y = Values)) +
  geom_point() + geom_line() +
  theme_bw()  + 
  ylab("Average Silhouette Width") + xlab("Number of clusters")

```

# Plot association group clusters at various levels
```{r, fig.width=12, fig.height=4}
for (k in c(3:10)) {
  dend.hel <- fviz_dend(assoc.clust, cex = 0.6, k = k,
                        color_labels_by_k = TRUE, hang = 10, 
                        k_colors = "d3", rect = TRUE, rect_border = "d3",
                        rect_fill = TRUE, main = paste(k,"groups"))
  print(dend.hel)
}

```

Extract group numbers and list members of each group
```{r}
A <- matrix(nrow = 160, ncol = 9)
A[,1] <- assoc.clust$labels
for (k in c(3:10)) {
  A[,k-1] <- cutree(assoc.clust, k = k)
}
A <- A[assoc.clust$order,]
colnames(A) <- c("Species","K3","K4","K5","K6","K7","K8","K9","K10")
assoc.groups <- as_tibble(A)

rm(A)
```

Merge indicator value analysis results
```{r}
assoc.groups <- assoc.groups %>% 
  left_join(indval.zoop.bioregions, by = "Species") %>% 
  left_join(indval.zoop.AB[,c(1:4,13)], by = "Species")

# write.csv(assoc.groups,
#           file = here::here("Data_Output/Assoc_indval_allcors_wardd2_11022021.csv"))
```

# Rotating the K = 5 dendrogram for plotting
```{r, fig.height = 6, fig.width = 15}
dend <- fviz_dend(assoc.clust, cex = 0.6, k = 5,
                      color_labels_by_k = TRUE, hang = 10, 
          k_colors = "d3", rect = TRUE, rect_border = "d3", rect_fill = TRUE,
          main = "K5")
dend

# Plotting only upper part of the dendrogram
dend <- assoc.clust %>% as.dendrogram
dend_h <- heights_per_k.dendrogram(dend)


od <- tibble(Species = labels(dend), Dend.Order = order.dendrogram(dend), 
             OG.taxa.order = c(1:160)) %>% 
  left_join(dplyr::select(assoc.groups, Species, K5)) %>% 
  mutate(AssocGroup = as.factor(K5)) 
od$AssocGroup <- fct_recode(od$AssocGroup,"ZO1" = "2", "ZO2" = "4",
                             "ZS1" = "1","ZS2"="3", "ZF" = "5")

neword <- data.frame(l2r.order = c(1:5), AssocGroup = unique(od$AssocGroup), 
                        New.l2r.order = c(2,1,5,3,4))
od <- left_join(od,neword) %>% arrange(New.l2r.order)
dend.r <- dendextend::rotate(dend, od$OG.taxa.order)


dend.hel <- fviz_dend(as.hclust(dend.r), cex = 0.4, k = 5, 
                      color_labels_by_k = TRUE, hang = 40, 
          k_colors = "d3", rect = TRUE, rect_border = "d3", rect_fill = TRUE,
          main = "K5-rotated")
dend.hel
```


# Kendall's W coefficient  of concordance on transformed species abundance
Global kendall test and a posteriori kendall test. The global test calculates the W and significance per group and the a posteriori test calculates the contribution of each species to its group W. The function documentation uses a hellinger transformation for the community data but here a log-chord transformation is used instead.
```{r, eval=FALSE}
set.seed(888)

# Get association grouping according to the order of column names in data table
asgrp <- od %>% arrange(Dend.Order) %>% dplyr::select(AssocGroup)
asgrp <- asgrp$AssocGroup
ken.w <- kendall.global(zoop.logchord, group = asgrp, mult="holm", nperm = 10000)
ken.w.post <- kendall.post(zoop.logchord, group = asgrp, mult="holm", nperm = 10000)
```

# Extract the kendall W test results to a dataframe
```{r, eval=FALSE}
# Global test
odnsp <-od %>% group_by(AssocGroup) %>% summarise(nsp = n())
ken.w.tab <- as.data.frame( ken.w$Concordance_analysis ) %>%  
  rownames_to_column(var = "Variable" ) %>% 
  pivot_longer( -Variable,
                names_to ="AssocGroup", values_to = "Values") %>% 
  mutate(AssocGroup = parse_number(AssocGroup)*10) %>% 
  mutate(AssocGroup = levels(asgrp)[AssocGroup]) %>% 
  pivot_wider(names_from = "Variable", values_from = "Values") %>% 
  left_join(odnsp) 
ken.w.tab$AssocGroup <- fct_relevel(ken.w.tab$AssocGroup,
                                    c("ZO1","ZO2","ZS1","ZS2","ZF"))
ken.w.tab <- arrange(ken.w.tab, AssocGroup)

# Posteriori test
ken.w.post.tab <- cbind.data.frame(ken.w.post$A_posteriori_tests_Group)
ken.w.post.tab <- as.data.frame( t(ken.w.post.tab) ) %>% 
  rownames_to_column(var = "Species") %>% 
  left_join(dplyr::select(od, Species, AssocGroup))

```

# Species association dendrogram
```{r}
# Set colors
clrs.assoc.k5 <- c("turquoise4","gold3","purple4","red4","olivedrab4")
swatch(clrs.assoc.k5)
clrs.assoc.k5sub <- c("turquoise4","turquoise3","gold3","gold",
                      "purple4","purple","red4","red1","olivedrab4")
swatch(clrs.assoc.k5sub)

# Updated dendrogram
dend.hel <- fviz_dend(as.hclust(dend.r), cex = 0.4, k = 5 , 
                      color_labels_by_k = TRUE, hang = 40, 
                      k_colors = clrs.assoc.k5, rect_border = clrs.assoc.k5, 
                      rect = TRUE, rect_fill = TRUE,
                      labels_track_height = 1.3,
                      main = "K5-rotated")
dend.hel

# png(here::here("Figures/Species_association_dendrogram_long.png"),
#     width = 11, height = 5, units = "in", res = 600)
# dend.hel
# dev.off()
```


# Table 3 of association groups by indval cluster
```{r}
# Load kendal test results
# load(here::here("Data_Output/Assoc_group_kendallresults_11022021.RData"))

# Categorize subgroups within each association group based on bioregion with the highest indicator value. For ZS2: OS, DS, and DF are grouped together as ZS2-Deep which all have indvals <25 anyway. The groups are then releveled to match the color coding.
sp.table <- indval.zoop.bioregions %>% 
  left_join(dplyr::select(ken.w.post.tab, -Prob), by = "Species") %>%
  mutate(AssocGroup = fct_relevel(AssocGroup, "ZO1","ZO2","ZS1","ZS2","ZF")) %>% 
  mutate(AssocGroup.sub = paste0(AssocGroup,"-",Bioregion)) %>% 
  mutate(AssocGroup.sub = fct_recode(AssocGroup.sub, 
                                     "ZO1-OS" = "ZO1-Offshore",
                                     "ZO1-DF" = "ZO1-DeepFjord",
                                     "ZO2-OS" = "ZO2-Offshore",
                                     "ZO2-NS" = "ZO2-Nearshore",
                                     "ZS1-DS" = "ZS1-DeepShelf",
                                     "ZS1-DF" = "ZS1-DeepFjord",
                                     "ZS2-NS" = "ZS2-Nearshore",
                                     "ZS2-Deep" = "ZS2-Offshore",
                                     "ZS2-Deep" = "ZS2-DeepShelf", 
                                     "ZS2-Deep" = "ZS2-DeepFjord",
                                     "ZF" = "ZF-DeepFjord")) %>% 
  mutate(AssocGroup.sub = fct_relevel(AssocGroup.sub,
                                      "ZO1-OS","ZO1-DF","ZO2-OS","ZO2-NS",
                                      "ZS1-DS","ZS1-DF","ZS2-NS","ZS2-Deep","ZF")) %>% 
  mutate(subclustnum = fct_recode(AssocGroup.sub, 
                                  "1" = "ZO1-OS", "2" = "ZO1-DF", "3" =  "ZO2-OS",
                                  "4" =  "ZO2-NS", "5" =  "ZS1-DS", "6" =  "ZS1-DF",
                                  "7" =  "ZS2-NS", "8" =  "ZS2-Deep", "9" =  "ZF")) %>% 
  mutate(subclustnum = as.numeric(subclustnum))


sp.table.output <- sp.table %>% 
  arrange(indval)  %>% 
  map_df(rev) %>% 
  group_by(AssocGroup.sub) %>% 
  summarise(nsp = n(), 
            alltypes = paste0(Species," (",format(round(indval, 1), nsmall = 1),")", collapse=", t"))

# write.csv(sp.table.output, 
#           file = here::here("Data_Output/Table_3_species_assoc_indval_11022021.csv"))
```


Plot rotated dendrogram with leaves colored by subgroup
```{r, fig.width=10, fig.height=5}
a <- data.frame(AssocGroup = levels(ken.w.tab$AssocGroup),
                AG.clr = clrs.assoc.k5)
b <- data.frame(AssocGroup.sub = levels(sp.table$AssocGroup.sub),
                AG.clr.sub = clrs.assoc.k5sub)
c <- as.hclust(dend.r)

spsubcolors <- data.frame(Species = c$labels[c$order],
                          order = c$order[c$order]) %>% 
  left_join(sp.table) %>% 
  left_join(a) %>% 
  left_join(b) 


dend.hel <- fviz_dend(as.hclust(dend.r), cex = 0.4, k = 5 , hang = 40, 
                      label_cols = spsubcolors$AG.clr.sub,
                      k_colors = clrs.assoc.k5, 
                      rect_border = clrs.assoc.k5, 
                      rect = FALSE, rect_fill = FALSE,
                      labels_track_height = 1.3,
                      main = "K5-rotated")
dend.hel

# png(here::here("Figures/Species_association_dendrogram_long_subassoc.png")),
#     width = 11, height = 5, units = "in", res = 600)
# dend.hel
# dev.off()
```

Figure 7a. Dendrogram with colored nodes according to bioregion with highest IndVal.
```{r, fig.width=8, fig.height=4}
d1 <- dend.r %>% 
  set("leaves_pch",19) %>% 
  set("leaves_cex",0.5) %>% 
  set("leaves_col",clrs.bioregion[spsubcolors$cluster]) %>% 
  set("branches_lwd",2)

d1 <- color_branches(d1, 5, col = clrs.assoc.k5)
d1 <- color_branches(d1, clusters = spsubcolors$subclustnum, 
                     col = clrs.assoc.k5sub[c(2,1,4,3,6,5,8,7,9)])
d1 %>% 
  set("labels",c()) %>% 
  plot()

# png(here::here("Figures/Species_association_dendrogram_colored_nodes_v2.png"),
#     width = 9, height = 4, units = "in", res = 600)
# d1 %>%
#   set("labels",c()) %>%
#   plot()
# dev.off()

```


# Supplementary table of assoc group and indval results
Columns are: Assoc group, nsp, (global W), subgroup, W per species, INDVAL analysis, % freq, mean abundance, sd abundance
```{r}
sp.table.output2 <- sp.table %>% 
  dplyr::select(-cluster,pval) %>% 
  left_join(dplyr::select(ken.w.tab, "AssocGroup","W"), by = "AssocGroup") %>% 
  left_join(indval.zoop.AB[,c(1:4,13)], by = "Species") %>% 
  left_join(sp.abund.list, by = "Species") %>% 
  arrange(indval)  %>% 
  map_df(rev) %>% 
  arrange(AssocGroup.sub)

# convert indval to percent
sp.table.output2[,12:15] <- sp.table.output2[,12:15]*100

# write.csv(sp.table.output2,
#           file = here::here("Data_Output/Supp_table_species_assoc_indval_11022021.csv"))

# # save kendal results
# save(file = here::here("Data_Output/Assoc_group_kendallresults_11022021.RData"),
#      ken.w.tab, ken.w.post.tab, clrs.assoc.k5, clrs.assoc.k5sub,
#      dend.hel, dend.r, assoc.clust, sp.table)
```

*RESTART HERE*
```{r}
# Load kendal test results
load(here::here("Data_Output/Assoc_group_kendallresults_11022021.RData"))
```


# NMDS of indicator value results with association group hulls
```{r, fig.height=11, fig.width=11}
set.seed(888)

indval.nmds <- metaMDS(indval.zoop.AB[,c(5:12)])
(indval.nmds)
stressplot(indval.nmds)
plot(indval.nmds, display = c("sites","species", choices = c(1,2)) )
points(indval.nmds, display = c("sites", "species"),
    choices = c(1,2)) 

# NMDS scores for biplot
indval.nmds.scores <- as.data.frame( scores(indval.nmds) ) %>% 
  rownames_to_column("Species") %>% 
  left_join(indval.zoop.bioregions) %>% 
  mutate(Bioregion = as.factor(Bioregion)) %>% 
  left_join(dplyr::select(sp.table, Species, AssocGroup, AssocGroup.sub)) %>% 
  ungroup()
indval.nmds.scores$Bioregion <- fct_relevel(indval.nmds.scores$Bioregion, 
                                            levels = c("Offshore","DeepShelf",
                                                       "Nearshore","DeepFjord"))
indval.nmds.scores$AssocGroup <- fct_relevel(indval.nmds.scores$AssocGroup,
                                    c("ZO1","ZO2","ZS1","ZS2","ZF"))

# Add to biplot the A & B components per cluster
AB.fit <- envfit(indval.nmds, indval.zoop.AB[,c(5:12)], 
                 permutations = 1000, strata = NULL,  choices=c(1,2), na.rm = TRUE)
AB.vec <- AB.fit$vectors$arrows * sqrt(AB.fit$vectors$r)
AB.vec <- as.data.frame(AB.vec) %>%  
  mutate(labs = c("SPC-OS","SPC-DS","SPC-NS","SPC-DF",
                  "FID-OS","FID-DS","FID-NS","FID-DF") )
```

Figure S7b. NMDS plots for association group level
```{r, fig.height=11, fig.width=11}

# Update sp names of some species
indval.nmds.scores$Species[indval.nmds.scores$Species == "Metridia aff lucens pseudopacifica"] <- "Metridia aff. lucens"
indval.nmds.scores$Species[indval.nmds.scores$Species == "Orthoconchoecia aff striola"] <- "Orthoconchoecia aff. striola"

# Identify species names to plot and font style
splabels <- filter(indval.nmds.scores, indval >= 25, padj < 0.05) %>% 
  left_join(dplyr::select(taxalist, Species = Taxa.Name, Taxa.Level)) %>% 
  mutate(Taxa.Level = if_else(Taxa.Level %in% c("C","O","F"),"plain","italic")) %>% 
  unique()

# Get convex hulls
find_hull <- function(df) df[chull(df$NMDS1, df$NMDS2), ]
hulls <- plyr::ddply(filter(indval.nmds.scores, indval >= 25, padj < 0.05),
                     "AssocGroup", find_hull)

g1 <- ggplot(indval.nmds.scores, 
       aes(x = NMDS1, y = NMDS2, color = AssocGroup)) +
  geom_point(size = 3)  +
  geom_polygon(data = hulls, aes(fill = AssocGroup), alpha=.1) +
  scale_color_manual(values = clrs.assoc.k5) +
  scale_fill_manual(values = clrs.assoc.k5) +
  theme_bw() +
  geom_text_repel(data = splabels, aes(label = Species, fontface = Taxa.Level),
                  size = 4, max.overlaps = 21) +
  geom_segment(data = AB.vec, aes(x = rep(0,nrow(AB.vec)), y =rep(0,nrow(AB.vec)),
                              xend = NMDS1,yend = NMDS2),
               lineend = "round", size = 1, colour = "black") +
  geom_text_repel(data = AB.vec, aes(x = NMDS1, y = NMDS2, label=labs), 
                  color="black", fontface = 'bold', size = 4) +
  theme(legend.position = "bottom",
        legend.title = element_text(color = 'black', size = 16, face = 'bold'),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 16))

# For association subgroup level
find_hull <- function(df) df[chull(df$NMDS1, df$NMDS2), ]
hulls <- plyr::ddply(filter(indval.nmds.scores, indval >= 25, padj < 0.05),
                     "AssocGroup.sub", find_hull) %>% 
  droplevels()
# correct colors for groups with no significant indval species
clrs.assoc.k5sub2 <- clrs.assoc.k5sub[which(levels(indval.nmds.scores$AssocGroup.sub)
                                            %in%
                                              levels(hulls$AssocGroup.sub))]

g2 <- ggplot(indval.nmds.scores, 
       aes(x = NMDS1, y = NMDS2, color = AssocGroup.sub)) +
  geom_segment(data = AB.vec, aes(x = rep(0,nrow(AB.vec)), y =rep(0,nrow(AB.vec)),
                              xend = NMDS1,yend = NMDS2),
               lineend = "round", size = 1, colour = "black") +
  geom_point(size = 3)  +
  geom_polygon(data = hulls, aes(fill = AssocGroup.sub), alpha=.1) +
  scale_color_manual(values = clrs.assoc.k5sub, name = "Association Group") +
  scale_fill_manual(values = clrs.assoc.k5sub2, guide = "none") +
  theme_bw() +
  geom_text_repel(data = splabels, aes(label = Species, fontface = Taxa.Level),
                  size = 4, max.overlaps = 21) +
  # geom_text_repel(data = AB.vec, aes(x = NMDS1, y = NMDS2, label=labs), 
  #                 color="black", fontface = 'bold', size = 4) +
  theme(legend.position = "bottom",
        legend.title = element_text(color = 'black', size = 16, face = 'bold'),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 16))

g1
g2

# png(here::here("Figures/Species_association_nmds_k5.png"),
#     width = 11, height = 11, units = "in", res = 600)
# g1
# dev.off()
# # Label arrows and revise legend in editor
png(here::here("Figures/Species_association_nmds_k5sub.png"),
    width = 11.5, height = 11.5, units = "in", res = 600)
g2
dev.off()
```

Figure 7 inset.
Indval nmds colored by bioregion assignment in indval. For subplot overlayed at upper rightside of assoc group nmds
```{r, fig.height=5, fig.width=5}
g3 <- ggplot(indval.nmds.scores, 
       aes(x = NMDS1, y = NMDS2, color = Bioregion)) +
  geom_point(size = 3) +
  scale_color_manual(values = clrs.bioregion) +
  theme_bw() +
  theme(legend.position = "none",
        panel.background = element_rect(fill = "transparent"), 
        plot.background = element_rect(fill = "transparent", color = NA),)

g3

png(here::here("Figures/Species_association_nmds_indval.png"),
    width = 3.5, height = 3, units = "in", res = 300)
g3
dev.off()
```

