---
title: "Robustness_analysis"
author: "Patrick Pata"
date: "30/10/2021"
output: html_document
---

Test the robustness of the k-means 4-cluster bioregionalization by randomly subsampling samples and species. Subsamples are compared to the overall bioregionalization and the % of samples within a subsample with different bioregion association are tallied.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load library and files with the original bioregionalization
```{r}
library(tidyverse)
library(vegan)
library(ape)
library(matrixStats)
library(here)
`%notin%` <- Negate(`%in%`)
source(here::here("Functions/p_biplot_pca.R"))

load(here::here("Data_Output/zoop_data_regionalized_10252021.RData"))

theme_set(theme_bw())
```

# Robustness to sampling methodology
1. Day vs night samples, even proportion of day and night samples (random subsampling)
2. Net types
3. Net mouth diameter
4. Mesh size
```{r}
set.seed(888)

# Day vs Night
zoopmeta %>% 
  group_by(Twilight) %>% 
  summarise(n = n())

keys <- zoopmeta %>% 
  dplyr::select(Twilight, Bioregion) %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))

for (t in c("Daylight","Night")) {
  ii <- which(keys$Twilight == t)
  sub.keys <- keys[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  

  print(paste0("Bioregionalization ",t," only: ",
               format(round(C, 2), nsmall = 2),
               "% classification error."))
  
  print(B/sum(A)*100)
}

# The same number of day and night samples (n = 773 each for 42% of samples)
bg.twilight <- matrix(nrow = 0, ncol = 8)
for (n in c(1:1000)){
  ii <- which(keys$Twilight == "Night")
  jj <- sample(filter(keys,Twilight == "Daylight")$idx, 
               size = 773, replace = FALSE)
  sub.keys <- keys[c(ii,jj),]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[c(ii,jj),]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  bg.twilight <- rbind(bg.twilight, c(n, nsub, nsp, C, (B/sum(A)*100)))
}

colnames(bg.twilight) <- c("n","nsub","nsp","percwrong","OS","DS","NS","DF")

bg.twilight %>% 
  as_tibble() %>% 
  summary()
```


```{r}
# Net types
zoopmeta %>% 
  group_by(Net.Type) %>% 
  summarise(n = n(), np = n()/3721*100)

zoopmeta %>% 
  group_by(Net.Type, Bioregion) %>% 
  summarise(n = n(), np = n()/3721*100)

keys <- zoopmeta %>% 
  dplyr::select(Net.Type, Bioregion) %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))


for (t in levels(zoopmeta$Net.Type) ) {
  ii <- which(keys$Net.Type == t)
  sub.keys <- keys[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  print(paste0("Bioregionalization ",t," only: ",
               format(round(C, 2), nsmall = 2),
               "% classification error."))
  print(B/sum(A)*100)
}

# *** non-bongo VNH ***
ii <- which(keys$Net.Type != "Bongo VNH")
sub.keys <- keys[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,3:4])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization non-Bongo VNH only: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))
 print(B/sum(A)*100)

# *** only SCOR and bongo VNH/ONH ***
ii <- which(keys$Net.Type %in% c("SCOR VNH","Bongo VNH","Bongo ONH") )
sub.keys <- keys[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,3:4])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization SCOR and bongo VNH/ONH only: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

print(B/sum(A)*100)


# The same number of net types samples (n = 175 at 35/type)
bg.nettype <- matrix(nrow = 0, ncol = 8)
for (n in c(1:1000)){
  ii <- which(keys$Net.Type == "Ring VNH")
  jj <- sample(filter(keys,Net.Type == "SCOR VNH")$idx,
               size = 35, replace = FALSE)
  kk <- sample(filter(keys,Net.Type == "Bongo VNH")$idx,
               size = 35, replace = FALSE)
  ll <- sample(filter(keys,Net.Type == "Bongo ONH")$idx,
               size = 35, replace = FALSE)
  mm <- sample(filter(keys,Net.Type == "NorPac VNH")$idx,
               size = 35, replace = FALSE)
  sub.keys <- keys[c(ii,jj,kk,ll,mm),]

  # recalculate log-chord table
  sub.abund <- zoop.log[c(ii,jj,kk,ll,mm),]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)

  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster

  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion.
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100

  bg.nettype <- rbind(bg.nettype, c(n, nsub, nsp, C, (B/sum(A)*100)))
}

colnames(bg.nettype) <- c("n","nsub","nsp","percwrong","OS","DS","NS","DF")

bg.nettype %>%
  as_tibble() %>%
  summary()
```


```{r}
# Net types
zoopmeta %>% 
  group_by(Net.Mouth) %>% 
  summarise(n = n(), np = n()/3721*100)

zoopmeta %>% 
  group_by(Net.Mouth, Bioregion) %>% 
  summarise(n = n(), np = n()/3721*100)

keys <- zoopmeta %>% 
  dplyr::select(Net.Mouth, Bioregion) %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))

for (t in unique(zoopmeta$Net.Mouth) ) {
  ii <- which(keys$Net.Mouth == t)
  sub.keys <- keys[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  print(paste0("Bioregionalization ",t," mouth diameter only: ",
               format(round(C, 2), nsmall = 2),
               "% classification error."))
  print(B/sum(A)*100)
}

# *** 0.56m and 0.58m only ***
ii <- which(keys$Net.Mouth == 0.56 | keys$Net.Mouth == 0.58)
# ii <- which(keys$Net.Mouth > 0.56)
sub.keys <- keys[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,3:4])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization 0.56-0.58m mouth diameter only: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

print(B/sum(A)*100)


# The same number of net mouth samples (n = 236 at 59/type)
bg.nettype <- matrix(nrow = 0, ncol = 8)
for (n in c(1:1000)){
  ii <- filter(keys,Net.Mouth == "0.42")$idx
  jj <- sample(filter(keys,Net.Mouth == 0.5)$idx,
               size = 59, replace = FALSE)
  kk <- sample(filter(keys,Net.Mouth == 0.56)$idx,
               size = 59, replace = FALSE)
  ll <- sample(filter(keys,Net.Mouth == 0.58)$idx,
               size = 59, replace = FALSE)
  sub.keys <- keys[c(ii,jj,kk,ll),]

  # recalculate log-chord table
  sub.abund <- zoop.log[c(ii,jj,kk,ll),]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)

  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster

  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion.
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100

  bg.nettype <- rbind(bg.nettype, c(n, nsub, nsp, C, (B/sum(A)*100)))
}

colnames(bg.nettype) <- c("n","nsub","nsp","percwrong","OS","DS","NS","DF")

bg.nettype %>%
  as_tibble() %>%
  summary()
```


```{r}
# Net types (group as 200-202, 236, 250-253, 330-335)
keys <- zoopmeta %>% 
  dplyr::select(Mesh.Size, Bioregion) %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))
keys$Mesh.Size[keys$Mesh.Size == 202] <- 200
keys$Mesh.Size[keys$Mesh.Size == 250] <- 253
keys$Mesh.Size[keys$Mesh.Size > 330] <- 330

keys %>% 
  group_by(Mesh.Size) %>% 
  summarise(n = n(), np = n()/3721*100)

keys %>% 
  group_by(Mesh.Size, Bioregion) %>% 
  summarise(n = n(), np = n()/3721*100)


for (t in unique(keys$Mesh.Size) ) {
  ii <- which(keys$Mesh.Size == t)
  sub.keys <- keys[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  print(paste0("Bioregionalization ",t," mouth diameter only: ",
               format(round(C, 2), nsmall = 2),
               "% classification error."))
  
  print(B/sum(A)*100)
}


# 236-253 um 
ii <- which(keys$Mesh.Size >= 236 & keys$Mesh.Size <= 253)
sub.keys <- keys[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,3:4])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization 236-253 um mouth diameter only: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

print(B/sum(A)*100)

# The same number of net mouth samples (n = 236 at 59/type)
bg.nettype <- matrix(nrow = 0, ncol = 8)
for (n in c(1:1000)){
  ii <- which(keys$Mesh.Size == 200)
  jj <- sample(filter(keys,Mesh.Size == 236)$idx,
               size = 48, replace = FALSE)
  kk <- sample(filter(keys,Mesh.Size == 253)$idx,
               size = 48, replace = FALSE)
  ll <- sample(filter(keys,Mesh.Size == 330)$idx,
               size = 48, replace = FALSE)
  sub.keys <- keys[c(ii,jj,kk,ll),]

  # recalculate log-chord table
  sub.abund <- zoop.log[c(ii,jj,kk,ll),]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)

  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster

  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion.
  A <- table(sub.keys[,3:4])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100

  bg.nettype <- rbind(bg.nettype, c(n, nsub, nsp, C, (B/sum(A)*100)))
}

colnames(bg.nettype) <- c("n","nsub","nsp","percwrong","OS","DS","NS","DF")

bg.nettype %>%
  as_tibble() %>%
  summary()
```


# Robustness to temporal extent
Number of samples per bioregion per year and per month
```{r}
A <- zoopmeta %>% 
  group_by(Bioregion, Year) %>% 
  summarise(n = n())
ggplot(A, aes(x = Year, y = n, color = Bioregion)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = clrs.bioregion)


B <- zoopmeta %>% 
  group_by(Bioregion, Month) %>% 
  summarise(n = n())
ggplot(B, aes(x = Month, y = n, color = Bioregion)) +
  geom_point() +
  geom_line() +
  scale_color_manual(values = clrs.bioregion)
```


# Robustness to timespan of analysis
Analyze by 5-year time chunk 
```{r}
set.seed(888)

keys <- bioregions %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))

bg.yearchunk <- matrix(nrow = 0, ncol = 8)


for (y in seq(1995,2014,5)) {
  ii <- which(zoopmeta$Year >= y & zoopmeta$Year <= y+4)
  sub.keys <- bioregions[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,2:3])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  print(C)
  print(B/sum(A)*100)

  # Save percentages
  bg.yearchunk <- rbind(bg.yearchunk, c(y, nsub, nsp, C, (B/sum(A)*100)))
  
}


colnames(bg.yearchunk) <- c("year","nsamples","nspecies","percwrong",
                            "OS","DS","NS","DF")
bg.yearchunk <- as_tibble(bg.yearchunk)
bg.yearchunk <- bg.yearchunk %>% 
  mutate(year = as.factor(year))


# png("Figures/SuppFig_robustness_year.png",
    # width = 7, height = 4, units = "in", res = 300)
ggplot(bg.yearchunk, aes(x = year, y = percwrong)) + 
  geom_point() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 10, linetype = "dotted") +
  xlab("Year") +
  ylab("Percent different bioregionalization")
# dev.off()
```

Analyze by 10-year time chunk 
```{r}
set.seed(888)

keys <- bioregions %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))

bg.yearchunk <- matrix(nrow = 0, ncol = 8)


for (y in seq(1995,2014,10)) {
  ii <- which(zoopmeta$Year >= y & zoopmeta$Year <= y+9)
  sub.keys <- bioregions[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,2:3])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  print(C)
  print(B/sum(A)*100)

  # Save percentages
  bg.yearchunk <- rbind(bg.yearchunk, c(y, nsub, nsp, C, (B/sum(A)*100)))
  
}


colnames(bg.yearchunk) <- c("year","nsamples","nspecies","percwrong",
                            "OS","DS","NS","DF")
bg.yearchunk <- as_tibble(bg.yearchunk)
bg.yearchunk <- bg.yearchunk %>% 
  mutate(year = as.factor(year))


ggplot(bg.yearchunk, aes(x = year, y = percwrong)) + 
  geom_point() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 10, linetype = "dotted") +
  xlab("Year") +
  ylab("Percent different bioregionalization")
```

1998-2014
```{r}
ii <- which(zoopmeta$Year >= 1998 & zoopmeta$Year <= 2014)
# ii <- which(zoopmeta$Year %in% seq(1995,2014,2))
sub.keys <- bioregions[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(C)
print(B/sum(A)*100)
```


# Analyze for each year
```{r}
set.seed(888)

bg.year <- matrix(nrow = 0, ncol = 8)

for (y in c(1995:2014)) {
  ii <- which(zoopmeta$Year == y)
  sub.keys <- bioregions[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,2:3])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  # Save percentages
  bg.year <- rbind(bg.year, c(y, nsub, nsp, C, 
                              (B/sum(A)*100)))
  
}


colnames(bg.year) <- c("year","nsamples","nspecies","percwrong",
                       "OS","DS","NS","DF")
bg.year <- as_tibble(bg.year)
bg.year <- bg.year %>% 
  mutate(year = as.factor(year))

bg.year2 <- bg.year %>% 
  pivot_longer(cols = c("OS","DS","NS","DF"), names_to = "Bioregion",
               values_to = "region_val") %>% 
  mutate(Bioregion = fct_relevel(as.factor(Bioregion),
                                 c("OS","DS","NS","DF")))

# png("Figures/Supp_figs/SuppFig_robustness_year.png",
# width = 7, height = 4, units = "in", res = 300)
ggplot(data = bg.year2, aes(x = year, y = region_val, 
                            color = Bioregion)) + 
  geom_point(alpha = 0.9) + 
  geom_point(data = bg.year, aes(x = year, y = percwrong), 
             color = "black") +
  scale_color_manual(values = clrs.bioregion)+
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 10, linetype = "dotted") +
  xlab("Year") +
  ylab("Classification error") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
# dev.off()

```

# Analyze per season
```{r}
ii <- which(zoopmeta$Month <= 6)
sub.keys <- bioregions[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100
print(paste("Apr-June:", C))


print(B/sum(A)*100)


ii <- which(zoopmeta$Month > 6 & zoopmeta$Month < 10)
sub.keys <- bioregions[ii,]

# recalculate log-chord table
sub.abund <- zoop.log[ii,]
sub.abund <- sub.abund[, colSums(sub.abund) > 0]
sub.abund <- decostand( sub.abund, method = "normalize")
nsub <- nrow(sub.abund)
nsp <- ncol(sub.abund)

# cluster with K-means, 4 bioregions
km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
sub.keys$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(sub.keys[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100
print(paste("July-Oct:", C))


print(B/sum(A)*100)
```


# Analyze for each month
```{r}
set.seed(888)

bg.month <- matrix(nrow = 0, ncol = 8)

for (m in c(4:10)) {
  ii <- which(zoopmeta$Month == m)
  sub.keys <- bioregions[ii,]
  
  # recalculate log-chord table
  sub.abund <- zoop.log[ii,]
  sub.abund <- sub.abund[, colSums(sub.abund) > 0]
  sub.abund <- decostand( sub.abund, method = "normalize")
  nsub <- nrow(sub.abund)
  nsp <- ncol(sub.abund)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
  sub.keys$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys[,2:3])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  # Save percentages
  bg.month <- rbind(bg.month, c(m, nsub, nsp, C, (B/sum(A)*100)))
}


colnames(bg.month) <- c("month","nsamples","nspecies","percwrong",
                        "OS","DS","NS","DF")
bg.month <- as_tibble(bg.month)
bg.month <- bg.month %>% 
  mutate(month = as.factor(month))


bg.month2 <- bg.month %>% 
  pivot_longer(cols = c("OS","DS","NS","DF"), names_to = "Bioregion",
               values_to = "region_val") %>% 
  mutate(Bioregion = fct_relevel(as.factor(Bioregion),
                                 c("OS","DS","NS","DF")))

# png(here::here("Figures/Supp_figs/SuppFig_robustness_month.png"),
# width = 7, height = 4, units = "in", res = 300)
ggplot(data = bg.month2, aes(x = month, y = region_val, 
                            color = Bioregion)) + 
  geom_point(alpha = 0.9) + 
  geom_point(data = bg.month, aes(x = month, y = percwrong), 
             color = "black") +
  scale_color_manual(values = clrs.bioregion)+
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 10, linetype = "dotted") +
  xlab("Month") +
  scale_x_discrete(breaks = c(4,5,6,7,8,9,10), 
                   labels = c("April","May","June","July",
                              "August","September","October")) +
  ylab("Classification Error")
# dev.off()



```


# Random subsampling with even number of sample per year (set at n=93 samples each year, 50% of all samples), will have to replace=TRUE subsampling for 1995 and 1996
```{r}
set.seed(888)
# nsub <- 93 # to get total 50% of all samples
nsub <- 78 # minimum number of samples in a year (42% of all samples)

keys <- bioregions %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))

bg.stratyear <- matrix(nrow = 0, ncol = 7)

for (n in c(1:1000)) {
  # Create new abundance matrix
  sub.abund.new <- zoop.log[0,]
  sub.keys.new <- bioregions[0,]
  
  for (y in c(1995:2014)) {
    ii <- which(zoopmeta$Year == y)
    sub.keys <- bioregions[ii,] %>% 
      rownames_to_column("idx") %>% 
      mutate(idx = as.numeric(idx))
    sub.abund <- zoop.log[ii,]
    
    # subsample rows
    # if (y <= 1996) {
    #   sk <- sample(keys$idx, size = nsub, replace = TRUE)
    # } else {
      sk <- sample(keys$idx, size = nsub, replace = FALSE)
    # }
    sub.keys <- keys[sk,]
    sub.abund <- zoop.log[sk,]
    
    sub.abund.new <- bind_rows(sub.abund.new, sub.abund)
    sub.keys.new <- bind_rows(sub.keys.new, sub.keys[,2:3])
  }
  
  # Bioregionalization
  
  # recalculate log-chord table
  sub.abund.new <- sub.abund.new[, colSums(sub.abund.new) > 0]
  sub.abund.new <- decostand( sub.abund.new, method = "normalize")
  nsp <- ncol(sub.abund.new)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund.new, 4, nstart = 10, iter.max = 1000)
  sub.keys.new$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys.new[,2:3])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  # Save percentages
  bg.stratyear <- rbind(bg.stratyear, c(nsp, n, C, (B/sum(A)*100) ))
}

colnames(bg.stratyear) <- c("nsp","n","percwrong","OS","DS","NS","DF")

bg.stratyear %>% 
  as_tibble() %>% 
  summary()

```

# Random subsampling by month (n = 266 for 50%) replace = TRUE for April
```{r}

set.seed(888)
# nsub <- 266 # to get total 50% of all samples
nsub <- 170 # minimum n samples per month (32%)

bg.stratmonth <- matrix(nrow = 0, ncol = 7)

for (n in c(1:1000)) {
  # Create new abundance matrix
  sub.abund.new <- zoop.log[0,]
  sub.keys.new <- bioregions[0,]
  
  for (m in c(4:10)) {
    ii <- which(zoopmeta$Month == m)
    sub.keys <- bioregions[ii,] %>% 
      rownames_to_column("idx") %>% 
      mutate(idx = as.numeric(idx))
    sub.abund <- zoop.log[ii,]
    
    # subsample rows
    # if (m == 4) {
    #   sk <- sample(keys$idx, size = nsub, replace = TRUE)
    # } else {
      sk <- sample(keys$idx, size = nsub, replace = FALSE)
    # }
    sub.keys <- keys[sk,]
    sub.abund <- zoop.log[sk,]
    
    sub.abund.new <- bind_rows(sub.abund.new, sub.abund)
    sub.keys.new <- bind_rows(sub.keys.new, sub.keys[,2:3])
  }
  
  # Bioregionalization
  
  # recalculate log-chord table
  sub.abund.new <- sub.abund.new[, colSums(sub.abund.new) > 0]
  sub.abund.new <- decostand( sub.abund.new, method = "normalize")
  nsp <- ncol(sub.abund.new)
  
  # cluster with K-means, 4 bioregions
  km <- kmeans(sub.abund.new, 4, nstart = 10, iter.max = 1000)
  sub.keys.new$newregions <- km$cluster
  
  # Get the % of samples with different bioregionalization
  # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
  A <- table(sub.keys.new[,2:3])
  B <- rowSums(A) - rowMaxs(A)
  C <- sum(B) / sum(A) * 100
  
  # Save percentages
  bg.stratmonth <- rbind(bg.stratmonth, c(nsp, n, C, (B/sum(A)*100)))
}

colnames(bg.stratmonth) <- c("nsp","n","percwrong","OS","DS","NS","DF")

bg.stratmonth %>% 
  as_tibble() %>% 
  summary()

```


# Subsampling net samples
levels: 5%:5%:95%, 99% of total number of samples
```{r}
set.seed(888)
subp <- c(seq(.05,.95,.05), .99)

keys <- bioregions %>% 
  rownames_to_column("idx") %>% 
  mutate(idx = as.numeric(idx))

bg.subsamp <- matrix(nrow = 0, ncol = 5)

for (i in subp) {
  for (n in c(1:1000)) { # 100 iterations for each subsample level
    nsub <- round( i * nrow(keys) )
    
    # subsample rows
    sk <- sample(keys$idx, size = nsub, replace = FALSE)
    sub.keys <- keys[sk,]
    
    # recalculate log-chord table
    sub.abund <- zoop.log[sk,]
    sub.abund <- sub.abund[, colSums(sub.abund) > 0]
    sub.abund <- decostand( sub.abund, method = "normalize")
    nsp <- ncol(sub.abund)
    
    # cluster with K-means, 4 bioregions
    km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
    sub.keys$newregions <- km$cluster
    
    # Get the % of samples with different bioregionalization
    # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
    A <- table(sub.keys[,3:4])
    B <- rowSums(A) - rowMaxs(A)
    C <- sum(B) / sum(A) * 100
    
    # Save percentages
    bg.subsamp <- rbind(bg.subsamp, c(i, nsub, nsp, n, C))
    
  }
}

colnames(bg.subsamp) <- c("subp","nsamples","nspecies","iternum","percwrong")
bg.subsamp <- as_tibble(bg.subsamp)
bg.subsamp <- bg.subsamp %>% 
  mutate(sperc = as.factor(subp * 100))


png(here::here("Figures/Supp_figs/SuppFig_robustness_samples_1000iter.png"),
width = 7, height = 4, units = "in", res = 300)
ggplot(bg.subsamp, aes(x = sperc, y = percwrong)) + 
  geom_boxplot() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 10, linetype = "dotted") +
  xlab("Percent of samples") +
  ylab("Classification error")
dev.off()

```



# Subsampling species
levels: 5%:5%:95%, 99% of total number of species

Note that kmeans can throw the error "Quick-TRANSfer stage steps exceeded maximum" which means it did not converge during that step. A solution is to increase nstart or use the Lloyd or MacQueen algorithms. 
(https://stackoverflow.com/questions/21382681/kmeans-quick-transfer-stage-steps-exceeded-maximum)
```{r}
set.seed(888)
subp <- c(seq(.05,.95,.05), .99)

bg.subspec <- matrix(nrow = 0, ncol = 5)
nsps <- c(1:ncol(zoop.log))

for (i in subp) {
  for (n in c(1:1000)) { # 100 iterations for each subsample level
    nsp <- round( i * length(nsps) )
    
    # subsample columns
    sk <- sample(nsps, size = nsp, replace = FALSE)
    
    # recalculate log-chord table
    sub.abund <- zoop.log[,sk]
    sub.abund <- sub.abund[rowSums(sub.abund) > 0,]
    sub.keys <- keys[which(rowSums(sub.abund) > 0), ] #update the subset of samples
    sub.abund <- decostand( sub.abund, method = "normalize")
    nsub <- nrow(sub.abund)
    
    # cluster with K-means, 4 bioregions
    km <- kmeans(sub.abund, 4, nstart = 10, iter.max = 1000)
    if (km$ifault==4) { 
      km = kmeans(sub.abund, 4, nstart = 20, iter.max = 1000,
                  algorithm="MacQueen")
      print(paste("Did not converge for:",i,n))
      }
    sub.keys$newregions <- km$cluster
    
    # Get the % of samples with different bioregionalization
    # Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
    A <- table(sub.keys[,3:4])
    B <- rowSums(A) - rowMaxs(A)
    C <- sum(B) / sum(A) * 100
    
    # Save percentages
    bg.subspec <- rbind(bg.subspec, c(i, nsub, nsp, n, C))
    
  }
}

colnames(bg.subspec) <- c("subp","nsamples","nspecies","iternum","percwrong")
bg.subspec <- as_tibble(bg.subspec)
bg.subspec <- bg.subspec %>% 
  mutate(sperc = as.factor(subp * 100))

png(here::here("Figures/Supp_figs/SuppFig_robustness_species_1000iter.png"),
    width = 7, height = 4, units = "in", res = 300)
ggplot(bg.subspec, aes(x = sperc, y = percwrong)) + 
  geom_boxplot() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_hline(yintercept = 10, linetype = "dotted") +
  xlab("Percent of species") +
  ylab("Classification error")
dev.off()

# ggplot(bg.subspec, aes(x = sperc, y = nsamples)) +
#   geom_boxplot() +
#   xlab("Percent of species") +
#   ylab("Number of samples")


```



