---
title: "Taxonomy_curation"
author: "Patrick Pata"
date: "18/10/2021"
output: html_document
---

This file loads the curated zooplankton dataset and extract the original taxonomic information. This will be manually curated in a spreadsheet. The curated spreadsheet is then loaded back for merging with the zooplankton dataset. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load files and libraries
```{r}
library(tidyverse)
library(here)
`%notin%` <- Negate(`%in%`)

load(here::here("Data_Output/zoop_curated_10202021.RData"))
theme_set(theme_bw())
```

# 1. Create species list for taxonomic curation
```{r, eval=FALSE}
taxa <- zoop_sub %>% 
  dplyr::select(Name, Phylum, Class, Order, Family, Genus, Species, Stage)
taxalist <- taxa %>% 
  unique()

# number of samples 
n1 <- length(unique(zoop_sub$Key))
n2 <- length(unique(filter(zoop_sub,Month >= 4 & Month <= 10)$Key))

# Count the number of samples a taxa was recorded for the entire dataset and also for the April-October period
A <- zoop_sub %>% 
  select(Key,Name) %>% unique() %>% count(Name) %>% 
  mutate(Name.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Name) %>%  unique() %>% count(Name) %>% 
  mutate(Name.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

A <- zoop_sub %>% 
  select(Key,Phylum) %>% unique() %>% count(Phylum) %>% 
  mutate(Phylum.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Phylum) %>%  unique() %>% count(Phylum) %>% 
  mutate(Phylum.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

A <- zoop_sub %>% 
  select(Key,Class) %>% unique() %>% count(Class) %>% 
  mutate(Class.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Class) %>%  unique() %>% count(Class) %>% 
  mutate(Class.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

A <- zoop_sub %>% 
  select(Key,Order) %>% unique() %>% count(Order) %>% 
  mutate(Order.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Order) %>%  unique() %>% count(Order) %>% 
  mutate(Order.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

A <- zoop_sub %>% 
  select(Key,Family) %>% unique() %>% count(Family) %>% 
  mutate(Family.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Family) %>%  unique() %>% count(Family) %>% 
  mutate(Family.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

A <- zoop_sub %>% 
  select(Key,Genus) %>% unique() %>% count(Genus) %>% 
  mutate(Genus.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Genus) %>%  unique() %>% count(Genus) %>% 
  mutate(Genus.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

A <- zoop_sub %>% 
  select(Key,Species) %>% unique() %>% count(Species) %>% 
  mutate(Species.Perc = floor(n/n1*10000)/100) %>% select(-n)
B <- zoop_sub %>% filter(Month >= 4 & Month <= 10) %>% 
  select(Key,Species) %>%  unique() %>% count(Species) %>% 
  mutate(Species.PercAO = floor(n/n2*10000)/100) %>% select(-n)
AB <- left_join(A,B)
taxalist <- left_join(taxalist, AB)

# Adjust order of columns
taxalist <- taxalist[,c(1,9,10,2,11,12,3,13,14,4,15,16,5,17,18,6,19,20,7,21,22,8)]
colnames(taxalist)

# Save to csv file for curation
# write.csv(taxalist, file = here::here("Data_Output/taxalist_10202021_unedited.csv"), row.names = FALSE)
```


# 2. Load and merge the updated taxalist to the zoop dataset
After this chunk, move to the next steps in the analysis. The updated data and taxalist is saved in the Data_Output folder.

The codes below are data exploration for the curated dataset.
```{r}
taxalist <- read.csv(here::here("Data_Raw/taxalist_curated_10252021.csv")) %>% 
  as_tibble() %>% 
  rename(Taxa.Name = "Updated.Name.2021",
         Taxa.Level = "Taxa.Level.2021",
         Taxa.Flag = "Flag.2021")

# Add taxa names if analyzed at Genus or Family level
# When analyzing at G/F level, update the flag and genus. Flag the taxa as 5 for instances when these were not analyzed at the species level but can be included when at the genus or family level.
taxalist <- taxalist %>% 
  mutate(Taxa.Name.Genus = Taxa.Name, 
         Taxa.Level.Genus = Taxa.Level, Taxa.Flag.Genus = Taxa.Flag,
         Taxa.Name.Family = Taxa.Name, 
         Taxa.Level.Family = Taxa.Level, Taxa.Flag.Family = Taxa.Flag)
# Genus level
ii <- which(taxalist$Taxa.Level == "S")
taxalist$Taxa.Name.Genus[ii] <- taxalist$Genus[ii]
taxalist$Taxa.Level.Genus[ii] <- "G"
jj <- which(taxalist$Taxa.Level == "S" & 
              (taxalist$Taxa.Flag == 2 | taxalist$Taxa.Flag == 3) )
taxalist$Taxa.Flag.Genus[jj] = 5

# Family level
ii <- which(taxalist$Taxa.Level %in% c("S","G"))
taxalist$Taxa.Name.Family[ii] <- taxalist$Family[ii]
taxalist$Taxa.Level.Family[ii] <- "F"
jj <- which(taxalist$Taxa.Level %in% c("S","G") & 
              (taxalist$Taxa.Flag == 2 | taxalist$Taxa.Flag == 3) )
taxalist$Taxa.Flag.Family[jj] = 5

zoop_sub <- zoop_sub %>% 
  left_join(dplyr::select(taxalist, Name, Taxa.Name, Taxa.Level,
                          Taxa.Flag, Major.Group,
                          Taxa.Name.Genus, Taxa.Level.Genus, Taxa.Flag.Genus,
                          Taxa.Name.Family, Taxa.Level.Family, Taxa.Flag.Family))

save("zoop_sub", "taxalist", file=here::here("Data_Output/zoop_curated_taxa_10252021.RData"))
```

