---
title: "Robustness_analysis_taxonomy"
author: "Patrick Pata"
date: "01/11/2021"
output: html_document
---

Robustness analysis to test similarity of results at genus and family level classification.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load library and files with the original bioregionalization
```{r}
ls.packages <- c("tidyverse", "vegan","matrixStats","here")
new.packages <- ls.packages[!(ls.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(tidyverse)
library(vegan)
library(here)
library(matrixStats)
`%notin%` <- Negate(`%in%`)
source(here::here("Functions/p_biplot_pca.R"))

load(here::here("Data_Output/zoop_data_regionalized_10252021.RData"))

theme_set(theme_bw())
```

# If not excluding species <3% presence
```{r}
set.seed(888)

# Create new cross tab
zs.sp.all <- zoop_sub %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.all[,1:2]
zs.sp <- zs.sp.all[,3:ncol(zs.sp.all)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization all 455 species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


# If using the same zoop_abundance table
Genus level
```{r}
set.seed(888)

A <- dplyr::select(taxalist, Taxa.Name, Taxa.Name.Genus) %>% 
  unique()

zs.genus <- zoop_sub %>% 
  left_join( A, by = c("Species" = "Taxa.Name" ) )

# Create new cross tab
zs.genus.abundance <- zs.genus %>% 
  group_by(Key, Bioregion, Genus = Taxa.Name.Genus) %>% 
  summarise(Abundance.m2 = sum(Abundance.m2),
            Biomass.mg.m2 = sum(Biomass.mg.m2)) %>% 
  ungroup() %>% 
  dplyr::select(Key, Bioregion, Genus, Abundance.m2) %>% 
  pivot_wider(names_from = Genus, values_from = Abundance.m2, values_fill = 0) 

# Original bioregionalization
keys.og <- zs.genus.abundance[,1:2]
zs.ga.ab <- zs.genus.abundance[,3:ncol(zs.genus.abundance)]

# Filter out <3% of all samples
zs.ga.pa <- decostand(zs.ga.ab, method = "pa")
ii <- which(colSums(zs.ga.pa / nrow(zs.ga.pa)) >= 0.03)
zs.ga.ab <- zs.ga.ab[,ii]

# transform
zs.ga.log <- log1p(zs.ga.ab)
zs.ga.logchord <- decostand( zs.ga.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.ga.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization at genus level: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


Family level
```{r}
set.seed(888)

A <- dplyr::select(taxalist, Taxa.Name, Taxa.Name.Family) %>% 
  unique()

zs.family <- zoop_sub %>% 
  left_join( A, by = c("Species" = "Taxa.Name" ) )

# Create new cross tab
zs.family.abundance <- zs.family %>% 
  group_by(Key, Bioregion, Family = Taxa.Name.Family) %>% 
  summarise(Abundance.m2 = sum(Abundance.m2),
            Biomass.mg.m2 = sum(Biomass.mg.m2)) %>% 
  ungroup() %>% 
  dplyr::select(Key, Bioregion, Family, Abundance.m2) %>% 
  pivot_wider(names_from = Family, values_from = Abundance.m2, values_fill = 0) 

# Original bioregionalization
keys.og <- zs.family.abundance[,1:2]
zs.fa.ab <- zs.family.abundance[,3:ncol(zs.family.abundance)]

# Filter out <3% of all samples
zs.fa.pa <- decostand(zs.fa.ab, method = "pa")
ii <- which(colSums(zs.fa.pa / nrow(zs.fa.pa)) >= 0.03)
zs.fa.ab <- zs.fa.ab[,ii]

# transform
zs.fa.log <- log1p(zs.fa.ab)
zs.fa.logchord <- decostand( zs.fa.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.fa.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization at family level: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100

```


# If including the observations that were excluded due to taxonomic specificity (flag = 2/3) 
Genus level
```{r}
set.seed(888)

A <- dplyr::select(taxalist, Taxa.Name, Taxa.Name.Genus) %>% 
  unique()

zs.genus <- zoop_sub %>% 
  left_join( A, by = c("Species" = "Taxa.Name" ) )

zs.genus.x <- zoop_sub_excluded %>% 
  left_join( A, by = c("Species" = "Taxa.Name" ) ) %>% 
  left_join( dplyr::select(zoopmeta, Key, Bioregion), by = "Key") %>% 
  filter(Taxa.Name.Genus %in% unique(zs.genus$Taxa.Name.Genus))


# Create new cross tab
zs.genus.abundance <- zs.genus %>% 
  bind_rows(zs.genus.x) %>% 
  group_by(Key, Bioregion, Genus = Taxa.Name.Genus) %>% 
  summarise(Abundance.m2 = sum(Abundance.m2),
            Biomass.mg.m2 = sum(Biomass.mg.m2)) %>% 
  ungroup() %>% 
  select(Key, Bioregion, Genus, Abundance.m2) %>% 
  pivot_wider(names_from = Genus, values_from = Abundance.m2, values_fill = 0) 

# Original bioregionalization
keys.og <- zs.genus.abundance[,1:2]
zs.ga.ab <- zs.genus.abundance[,3:ncol(zs.genus.abundance)]

# Filter out <3% of all samples
zs.ga.pa <- decostand(zs.ga.ab, method = "pa")
ii <- which(colSums(zs.ga.pa / nrow(zs.ga.pa)) >= 0.03)
zs.ga.ab <- zs.ga.ab[,ii]

# transform
zs.ga.log <- log1p(zs.ga.ab)
zs.ga.logchord <- decostand( zs.ga.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.ga.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization at genus level: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```

Family level
```{r}
set.seed(888)

A <- dplyr::select(taxalist, Taxa.Name, Taxa.Name.Family) %>% 
  unique()

zs.family <- zoop_sub %>% 
  left_join( A, by = c("Species" = "Taxa.Name" ) )

zs.family.x <- zoop_sub_excluded %>% 
  left_join( A, by = c("Species" = "Taxa.Name" ) ) %>% 
  left_join( dplyr::select(zoopmeta, Key, Bioregion), by = "Key") %>% 
  filter(Taxa.Name.Family %in% unique(zs.family$Taxa.Name.Family))


# Create new cross tab
zs.family.abundance <- zs.family %>% 
  bind_rows(zs.family.x) %>% 
  group_by(Key, Bioregion, Family = Taxa.Name.Family) %>% 
  summarise(Abundance.m2 = sum(Abundance.m2),
            Biomass.mg.m2 = sum(Biomass.mg.m2)) %>% 
  ungroup() %>% 
  dplyr::select(Key, Bioregion, Family, Abundance.m2) %>% 
  pivot_wider(names_from = Family, values_from = Abundance.m2, values_fill = 0) 

# Original bioregionalization
keys.og <- zs.family.abundance[,1:2]
zs.fa.ab <- zs.family.abundance[,3:ncol(zs.family.abundance)]

# Filter out <3% of all samples
zs.fa.pa <- decostand(zs.fa.ab, method = "pa")
ii <- which(colSums(zs.fa.pa / nrow(zs.fa.pa)) >= 0.03)
zs.fa.ab <- zs.fa.ab[,ii]

# transform
zs.fa.log <- log1p(zs.fa.ab)
zs.fa.logchord <- decostand( zs.fa.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.fa.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization at family level: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100

```

Copepods only
```{r}
set.seed(888)

# Create subset species list
# splist <- left_join(sp.abund.list, taxalist, by = c("Species" = "Taxa.Name"))
splist <- taxalist %>% 
  filter(Major.Group %in% c("Calanoids","Non-calanoids") & Taxa.Flag == 1) %>% 
  dplyr::select(Species = "Taxa.Name", Major.Group) %>% 
  distinct() %>% 
  left_join(sp.abund.list) %>% 
  filter(presence.perc >= 3)

# Create new cross tab
zs.sp.cop <- zoop_sub %>% 
  filter(Species %in% splist$Species) %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.cop[,1:2]
zs.sp <- zs.sp.cop[,3:ncol(zs.sp.cop)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization for ",nrow(splist)," copepod species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```

Holoplankton
```{r}
# Create subset species list
# splist <- left_join(sp.abund.list, taxalist, by = c("Species" = "Taxa.Name"))
splist <- taxalist %>% 
  filter(Major.Group %in% c("Calanoids","Non-calanoids","Appendicularians",
                            "Branchiopods","Gastropods","Hydromedusae",
                            "Siphonophores","Amphipods","Mysids","Euphausiids",
                            "Chaetognaths","Ctenophora","Ostracods","Salps",
                            "Doliolids","Annelids") & Taxa.Flag == 1) %>% 
  filter(Order %notin% c("Canalipalpata","Neotaenioglossa")) %>% 
  filter(Taxa.Flag == 1) %>% 
  dplyr::select(Species = "Taxa.Name", Major.Group) %>% #, Family, Order, Class) %>% 
  distinct() %>% 
  left_join(sp.abund.list) %>% 
  filter(presence.perc >= 3)

# Create new cross tab
zs.sp.cop <- zoop_sub %>% 
  filter(Species %in% splist$Species) %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.cop[,1:2]
zs.sp <- zs.sp.cop[,3:ncol(zs.sp.cop)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization for ",nrow(splist)," holoplankton species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


Meroplankton
```{r}
# Create subset species list
# splist <- left_join(sp.abund.list, taxalist, by = c("Species" = "Taxa.Name"))
splist <- taxalist %>% 
  filter(Major.Group %notin% c("Calanoids","Non-calanoids","Appendicularians",
                            "Branchiopods","Hydromedusae", "Amphipods",
                            "Siphonophores","Mysids","Euphausiids",
                            "Chaetognaths","Ctenophora","Ostracods","Salps",
                            "Doliolids") & Taxa.Flag == 1) %>%
  filter(Order %notin% c("Aciculata","Gymnosomata","Thecosomata","Cephalaspidea")) %>%
  dplyr::select(Species = "Taxa.Name", Major.Group) %>%  #, Family, Order, Class) %>% 
  distinct() %>% 
  left_join(sp.abund.list) %>%
  filter(presence.perc >= 3)

# Create new cross tab
zs.sp.cop <- zoop_sub %>% 
  filter(Species %in% splist$Species) %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.cop[,1:2]
zs.sp <- zs.sp.cop[,3:ncol(zs.sp.cop)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization for ",nrow(splist)," meroplankton species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


Crustacean Holoplankton
```{r}
# Create subset species list
# splist <- left_join(sp.abund.list, taxalist, by = c("Species" = "Taxa.Name"))
splist <- taxalist %>% 
  filter(Major.Group %in% c("Calanoids","Non-calanoids",
                            "Branchiopods","Amphipods","Mysids","Euphausiids",
                            "Ostracods") & Taxa.Flag == 1) %>% 
  filter(Order %notin% c("Canalipalpata")) %>% 
  filter(Taxa.Flag == 1) %>% 
  dplyr::select(Species = "Taxa.Name", Major.Group) %>% #, Family, Order, Class) %>% 
  distinct() %>% 
  left_join(sp.abund.list) %>% 
  filter(presence.perc >= 0)

# Create new cross tab
zs.sp.cop <- zoop_sub %>% 
  filter(Species %in% splist$Species) %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.cop[,1:2]
zs.sp <- zs.sp.cop[,3:ncol(zs.sp.cop)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization for ",nrow(splist)," crustacean holoplankton species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


non-crustacean Holoplankton
```{r}
# Create subset species list
# splist <- left_join(sp.abund.list, taxalist, by = c("Species" = "Taxa.Name"))
splist <- taxalist %>% 
  filter(Major.Group %in% c("Gastropods","Hydromedusae",
                            "Siphonophores", "Appendicularians",
                            "Chaetognaths","Ctenophora","Salps",
                            "Doliolids","Annelids") & Taxa.Flag == 1) %>% 
  filter(Order %notin% c("Canalipalpata","Neotaenioglossa")) %>% 
  filter(Taxa.Flag == 1) %>% 
  dplyr::select(Species = "Taxa.Name", Major.Group) %>% #, Family, Order, Class) %>% 
  distinct() %>% 
  left_join(sp.abund.list) %>% 
  filter(presence.perc >= 0)

# Create new cross tab
zs.sp.cop <- zoop_sub %>% 
  filter(Species %in% splist$Species) %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.cop[,1:2]
zs.sp <- zs.sp.cop[,3:ncol(zs.sp.cop)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization for ",nrow(splist)," non-crustacean holoplankton species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


Holoplankton w/o copepods
```{r}
# Create subset species list
# splist <- left_join(sp.abund.list, taxalist, by = c("Species" = "Taxa.Name"))
splist <- taxalist %>% 
  filter(Major.Group %in% c("Appendicularians",
                            "Branchiopods","Gastropods","Hydromedusae",
                            "Siphonophores","Amphipods","Mysids","Euphausiids",
                            "Chaetognaths","Ctenophora","Ostracods","Salps",
                            "Doliolids","Annelids") & Taxa.Flag == 1) %>% 
  filter(Order %notin% c("Canalipalpata","Neotaenioglossa")) %>% 
  filter(Taxa.Flag == 1) %>% 
  dplyr::select(Species = "Taxa.Name", Major.Group) %>% #, Family, Order, Class) %>% 
  distinct() %>% 
  left_join(sp.abund.list) %>% 
  filter(presence.perc >= 3)

# Create new cross tab
zs.sp.cop <- zoop_sub %>% 
  filter(Species %in% splist$Species) %>% 
  dplyr::select(Key, Bioregion, Species, Abundance.m2) %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.cop[,1:2]
zs.sp <- zs.sp.cop[,3:ncol(zs.sp.cop)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization for ",nrow(splist)," holoplankton species: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```


# Bioregionalization of the presence-absence and biomass data
```{r}
set.seed(888)

# ** P/A **
# Original bioregionalization
keys.og <- bioregions

# Chord transformation of presence-absence
zs.sp.chord <- decostand( zoop.pa, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zoop.pa, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# # Sorensen dissimilarity
# zs.sp.soren <- vegdist(zoop.pa, method = "bray", binary = TRUE)
# km <- kmeans(zs.sp.soren, 4, nstart = 10, iter.max = 1000)

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization of presence-absence: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100



# ** Biomass **
splist <- sp.abund.list %>% 
  filter(presence.perc >= 3)

# Create new cross tab
zs.sp.biom <- zoop_sub %>% 
  dplyr::select(Key, Bioregion, Species, Biomass.mg.m2) %>% 
  filter(Species %in% splist$Species) %>% 
  pivot_wider(names_from = Species, values_from = Biomass.mg.m2, values_fill = 0) 


# Original bioregionalization
keys.og <- zs.sp.biom[,1:2]
zs.sp <- zs.sp.biom[,3:ncol(zs.sp.biom)]


# transform
zs.sp.log <- log1p(zs.sp)
zs.sp.logchord <- decostand( zs.sp.log, method = "normalize")

# cluster with K-means, 4 bioregions
km <- kmeans(zs.sp.logchord, 4, nstart = 10, iter.max = 1000)
keys.og$newregions <- km$cluster

# Get the % of samples with different bioregionalization
# Consider that the maximum cluster for each row in the crosstab table corresponds to the original bioreigion. 
A <- table(keys.og[,2:3])
B <- rowSums(A) - rowMaxs(A)
C <- sum(B) / sum(A) * 100

print(paste0("Bioregionalization of biomass: ",
             format(round(C, 2), nsmall = 2),
             "% classification error."))

B/sum(A)*100
```

