---
title: "Variance_partitioning"
author: "Patrick Pata"
date: "02/11/2021"
output: html_document
---
Creats figure 5 of the Pata et al. PinO manuscript.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load files and libraries
```{r}
ls.packages <- c("tidyverse", "vegan","ggrepel","tictoc","here")
new.packages <- ls.packages[!(ls.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)

library(tidyverse)
library(vegan)
library(ggrepel)
library(tictoc)
library(here)
`%notin%` <- Negate(`%in%`)

load(here::here("Data_Output/zoop_data_regionalized_10252021.RData"))
load(here::here("Data_Output/dbMEM_significant_MEMS_11072021.RData"))
```

Variance partitioning of spatial dbMEM and all environmental variables
```{r}
vars.envi2 <- select(vars.envi, -Key)

dbmem.broad <- bind_cols(vars.xy, dbmem.space.broad)
dbmem.fine <- dbmem.space.medfine


# Plot as dummy variable for variance partitioning
breg <- as.data.frame(model.matrix(~ zoopmeta$Bioregion)[,-1])
rda.reg <- rda(zoop.logchord, breg)
sum(rda.reg$CCA$eig)/rda.reg$tot.chi*100

# For subset with RDA rows only
a <- left_join(vars.envi, bioregions)
breg <- as.data.frame(model.matrix(~ a$Bioregion)[,-1])
rda.reg <- rda(zoop.rda, breg)
sum(rda.reg$CCA$eig)/rda.reg$tot.chi*100

varpart(zoop.rda, vars.xy, dbmem.space.broad)
varpart(zoop.rda, dbmem.space.broad, dbmem.space.medfine)


vp <- varpart(zoop.rda, vars.envi2,
              bind_cols(vars.xy, dbmem.broad), 
              breg)
vp
plot(vp, digits = 2, Xnames = c("Environment","Broad Space","Bioregions"))

```

Variance explained by methodological differences
```{r}
vars.meta <- zoop.rda %>% 
  rownames_to_column("Key") %>% 
  select(Key) %>% 
  left_join(select(zoopmeta, Key, Twilight, Tow.Depth, Volume.Filtered.m3,
                   Tow.Ratio, Net.Type, Net.Mouth, Mesh.Size, Time)) %>% 
  mutate(Net.Mouth = as.factor(Net.Mouth),
         Mesh.Size = as.factor(Mesh.Size))

rda.meta <- rda(zoop.rda ~ Twilight + Net.Mouth + Mesh.Size + Net.Type,
                data = vars.meta)
(RsquareAdj(rda.meta)$adj.r.squared)
```

Figure 5a. RDA with 4 environmental variables.
```{r}
# RDA
vars.envi3 <- vars.envi2 %>% 
  select(depth, salinity, temperature, microplankton)
rda.all <- rda(zoop.rda ~ ., vars.envi3)
(R2adj <- RsquareAdj(rda.all)$adj.r.squared)
rda.all.ax12 <- scores(rda.all, choices = c(1:2), display = "lc")

# variance explained of the constrained axes
varexp.RDA <- round( rda.all$CCA$eig/rda.all$tot.chi*100, digits = 1)
head(varexp.RDA)    

# variance explained of the unconstrained axes
varexp.PC <- round( rda.all$CA$eig/rda.all$tot.chi*100, digits = 1)
head(varexp.PC)    

# Scaling 2 is a correlation triplot... angles between arrows reflect correlations.
rda.scores.arrows <- scores(rda.all, choices = c(1:4), display = "bp", scaling = 2) 
rda.scores.sites <- scores(rda.all, choices = c(1:4), display = "lc", scaling = 2) 

rda.scores.sites <- as.data.frame(rda.scores.sites) %>% 
  rownames_to_column("Key") %>% 
  left_join(bioregions)

rda.arrow.df <- data.frame(variable = rownames(rda.scores.arrows),
                           x1 = rep(0,nrow(rda.scores.arrows)),
                           y1 =rep(0,nrow(rda.scores.arrows)), 
                           xend = rda.scores.arrows[,1],
                           yend = rda.scores.arrows[,2])
      
fig5a <- ggplot(rda.scores.sites, 
       aes_string("RDA1", "RDA2", colour = "Bioregion")) + 
  geom_point(size = 3, alpha = 0.3, shape = 16) + 
  stat_ellipse(size = 1.5,show.legend = FALSE,level = 0.9) +
  geom_vline(xintercept = 0, colour='darkgray') + 
  geom_hline(yintercept = 0, colour='darkgray') +
  theme_bw()+
  theme(legend.position="none", 
        legend.background = element_rect(colour="gray", 
                                         size=0.5, linetype="solid"),
        legend.title = element_text(color = 'black', size = 16, face = 'bold'),
        legend.text = element_text(size = 16),
        axis.title = element_text(size = 16),
        axis.text = element_text(size = 16)) + 
  labs(x = paste0("RDA1 (",varexp.RDA[1],"%)"), 
       y = paste0("RDA2 (",varexp.RDA[2], "%)" )) +
  scale_color_manual(values = clrs.bioregion, name = "Bioregion") +
  geom_segment(data = rda.arrow.df, aes(x = x1, y = y1, 
                                        xend = xend, yend = yend),
               lineend = "round", size = 1, colour = "black",
               arrow = arrow(length = unit(0.21, "cm"))) +
  geom_text_repel(data = rda.arrow.df, aes(x = xend, y = yend,
                                           label=variable), 
                  color="black", size = 5) +
  scale_x_reverse()



pdf(here::here("Figures/Figure_5a.pdf"),
    width = 6, height = 6)
fig5a
dev.off()
```

Figure 5b. Variance partitioning with 4 environmental variables. This figure is manually edited to add color and adjust layout.
```{r}
vp <- varpart(zoop.rda, vars.envi3,
              bind_cols(vars.xy, dbmem.space.broad), 
              breg)
vp
plot(vp, digits = 2, Xnames = c("Environment","Broad Space","Bioregions"))


```
