---
title: "Bioregions_geographic_space"
author: "Patrick Pata"
date: "21/10/2021"
output: html_document
---

First individual maps of the bioregions are generated and combined maps which may have overlapping points. The overlap in points can be organized by gridding the area and selecting the most common bioregoin assignment for that area.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load files and libraries
```{r}
library(tidyverse)
library(vegan)
library(ape)
library(tictoc)
library(rnaturalearth)
library(marmap)
library(mapdata)
library(egg)
library(cowplot)
library(here)

`%notin%` <- Negate(`%in%`)

load(here::here("Data_Output/zoop_data_regionalized_10252021.RData"))

# source(here::here("Functions/p_biplot_pca.R"))


```

# Setup the background map and bathymetric contours
```{r}
world <- ne_countries(scale = "large", returnclass = "sf")

# get bathymetry data
bathy = getNOAA.bathy(lon1 = -150, lon2 = -122, lat1 = 45, lat2 = 60, resolution = 1)
# convert bathymetry to data frame
bathy = fortify.bathy(bathy)
```


Fig 1. Distribution of samples in space and time
```{r, fig.width=10, fig.height=9.5}
zoopmeta2 <- zoopmeta %>% 
  arrange(Year) %>% 
  mutate(`Time of day` = Twilight) %>% 
  mutate(`Time of day` = fct_recode(`Time of day`,"Day" = "Daylight"))

g1 <- ggplot(data = world) +
  geom_sf(fill = "#A9A9A9") + 
  coord_sf(xlim = c(-145.25, -122), ylim = c(47, 56.1), expand = FALSE) +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200,-500,-1000), size=c(0.2), colour="black") +
  xlab("") + ylab("") + theme_bw() + 
  theme(legend.position =  "bottom", 
        legend.key.width = unit(1, "inch"),
        legend.title = element_text(color = 'black', size = 12),
        legend.text = element_text(size = 12), 
        plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12), 
        axis.text = element_text(size = 12)) +
  scale_x_continuous(breaks = c(-145,-140,-135,-130,-125)) +
  geom_point(data = zoopmeta2, 
             aes(x = Longitude, y = Latitude, colour = Year), 
            size = 1.5, alpha = 0.9) +
  scale_color_viridis_c(breaks = seq(1996, 2014, 2), expand = c(0, 0), 
                        name = "Sample\nYear")

g2 <- ggplot(data = zoopmeta2,
       aes(x = Year, fill = `Time of day`)) +
  geom_bar(position = "stack") +
  ylab("# Samples") +
  scale_x_continuous(breaks = c(1995,2000,2005,2010))+ 
  scale_y_continuous(expand = c(0,0)) +
  theme(axis.title = element_text(size = 12), 
        axis.text = element_text(size = 12), 
        legend.position = "none") 
g3 <- ggplot(data = zoopmeta2,
       aes(x = as.factor(Month), fill = `Time of day`)) +
  geom_bar(position = "stack") +
  ylab("# Samples") + xlab("Month") +
  scale_x_discrete(labels = c("Apr","May","Jun","Jul","Aug","Sep","Oct")) +
  scale_y_continuous(expand = c(0,0)) +
  theme(legend.title = element_text(color = 'black', size = 12),
        legend.text = element_text(size = 12), 
        axis.title = element_text(size = 12), 
        axis.text = element_text(size = 12)) 

g4 <- cowplot::plot_grid(g2,g3, align = "h", nrow = 1, 
                         rel_widths = c(0.85,1))

png(here::here("Figures/Fig1_sampledist_v3.png"),
width = 10, height = 9.5, units = "in", res = 600)
cowplot::plot_grid(g1,g4, ncol = 1, rel_heights = c(3,1))
dev.off()
```


Create the ggplot object which is the background map
```{r, fig.width=10, fig.height=7}
map.bc <- ggplot(data = world) +
  geom_sf(fill = "#A9A9A9") + 
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200,-500,-1000), size=c(0.2), colour="black") +
  # geom_contour(data = bathy, aes(x=x, y=y, z=z),
  #              breaks=c(-200), size=c(0.6), colour="grey") +
  coord_sf(xlim = c(-145.25, -122), ylim = c(47, 56.1), expand = FALSE) +
  xlab("") + ylab("") + theme_bw() + 
  theme(legend.position =  "none", 
        legend.title = element_text(color = 'black', size = 16),
        legend.text = element_text(size = 16), 
        plot.title = element_text(size = 16, face = "bold"),
        axis.title = element_text(size = 16), 
        axis.text = element_text(size = 16)) +
  scale_x_continuous(breaks = c(-145,-140,-135,-130,-125))

map.bc
```

All bioregions
```{r, fig.width=10, fig.height=7}
g <- map.bc +
  geom_point(data = zoopmeta, 
             aes(x = Longitude, y = Latitude, colour = Bioregion), 
            size = 2, alpha = 0.7) +
  scale_color_manual(values = clrs.bioregion)

g
# png(here::here("Figures/Bioregionalization_map_points.png"),
#     width = 10, height = 7, units = "in", res = 600)
# g
# dev.off()
```


```{r, fig.width=11, fig.height=7}
g1 <- map.bc +
  geom_point(data = filter(zoopmeta, Bioregion %in% c("Offshore")), 
             aes(x = Longitude, y = Latitude, colour = Bioregion), 
            size = 3, alpha = 0.7, shape = 18) +
  geom_point(data = filter(zoopmeta, Bioregion %in% c("Nearshore")), 
             aes(x = Longitude, y = Latitude, colour = Bioregion), 
            size = 2, alpha = 0.7, shape = 16) +
  scale_color_manual(values = clrs.bioregion[c(3,1)])

g2 <- map.bc +
  geom_point(data = filter(zoopmeta, Bioregion %in% c("DeepShelf")), 
             aes(x = Longitude, y = Latitude, colour = Bioregion), 
            size = 3, alpha = 0.7, shape = 18) +
  geom_point(data = filter(zoopmeta, Bioregion %in% c("DeepFjord")), 
             aes(x = Longitude, y = Latitude, colour = Bioregion), 
            size = 2, alpha = 0.7, shape = 16) +
  scale_color_manual(values = clrs.bioregion[c(4,2)]) +
  coord_sf(xlim = c(-136, -122), ylim = c(47, 56.1), expand = FALSE) +
  theme(axis.text.y = element_blank())

g1
g2

# png(here::here("Figures/Bioregionalization_map_points_A.png"),
#     width = 10, height = 7, units = "in", res = 600)
# print(g1)
# dev.off()
# 
# png(here::here("Figures/Bioregionalization_map_points_B.png"),
#     width = 10, height = 7, units = "in", res = 600)
# print(g2)
# dev.off()

# width of map1 = 23.25, map2 = 14, proportion = 1.66:1
# png(here::here("Figures/Bioregionalization_map_points_AB.png"),
#     width = 11, height = 7, units = "in", res = 900)
# plot_grid(g1,g2, align = "h", axis = "bl", ncol = 2, rel_widths = c(1.66,1))
# dev.off()
```


Grid according to most commonly represent bioregion.
```{r, fig.width=10, fig.height=7}
res <- 0.25
lon_grid <- seq(-145.25,-122,res)
lat_grid <- seq(47, 56, res)

bioregions.gridded <- zoopmeta %>% 
  dplyr::select(Key,Bioregion,Longitude,Latitude) %>% 
  mutate(lon_bin = lon_grid[ .bincode(Longitude, breaks = lon_grid) ], 
         lat_bin = lat_grid[ .bincode(Latitude, breaks = lat_grid) ]) %>% 
  group_by(Bioregion, lon_bin, lat_bin) %>% 
  summarise(n = n()) %>% 
  group_by(lon_bin, lat_bin) %>% 
  top_n(n=1) %>% 
  arrange(Bioregion)

# duplicated grid points
bioregions.gridded$isdistinct <- duplicated(dplyr::select(bioregions.gridded, lon_bin, lat_bin))

# Create triangle polygons for the duplicated grid cells
dups <- bioregions.gridded %>% 
  ungroup() %>% 
  filter(isdistinct == TRUE) %>% 
  rownames_to_column(var = "index") %>% 
  mutate(x1 = lon_bin, x2 = lon_bin+res, x3 = lon_bin+res,
         y1 = lat_bin, y2 = lat_bin, y3 = lat_bin+res) %>% 
  dplyr::select(-c(isdistinct,n,lon_bin,lat_bin))
A <- dplyr::select( pivot_longer(dplyr::select(dups, index, Bioregion, x1, x2, x3), 
                          cols = c(x1, x2, x3), names_to = "x.edge", values_to = "x"),
             -x.edge)
B <- dplyr::select( pivot_longer(dplyr::select(dups, index, Bioregion, y1, y2, y3), 
                          cols = c(y1, y2, y3), names_to = "y.edge", values_to = "y"), 
             -y.edge)
dups <- cbind(A,y = B$y)
rm(A, B)


g <- map.bc +
  geom_tile(data = filter(bioregions.gridded, isdistinct == FALSE),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Bioregion),
            alpha = 0.8) +
  geom_polygon(data = dups, aes(x = x, y = y, group = index, fill=Bioregion)) +
  scale_fill_manual(values = clrs.bioregion) + 
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500,-1000), size=c(0.2), colour="black") 

g
# png(here::here("Figures/Bioregionalization_map_gridded.png"),
#     width = 10, height = 7, units = "in", res = 600)
# g
# dev.off()

```



